---
title: "Disability Status and Oral Health Outcomes"
output: html_notebook
---

For this project, I will be using the National Health and Nutrition Examination Survey (NHANES) data (2017-2018).

```{r}
library(haven)
library(tidyverse)
library(ggplot2)
library(ggmosaic)
library(survey)
library(esquisse)
library(survey)
```

Reading in NHANES 2017-2018 data.
```{r}
# Reading in "Demographic Variables and Sample Weights" data (covariates)
demo <- read_xpt("https://wwwn.cdc.gov/Nchs/Nhanes/2017-2018/DEMO_J.XPT") %>%
  select(SEQN, RIDSTATR, RIAGENDR, RIDAGEYR, RIDRETH3, DMDEDUC2,
  WTINT2YR, WTMEC2YR, SDMVPSU, SDMVSTRA)
  
# Reading in health insurance data (covariate)
insure <- read_xpt("https://wwwn.cdc.gov/Nchs/Nhanes/2017-2018/HIQ_J.XPT") %>%
  select(SEQN, HIQ011, HIQ031A, HIQ031B, HIQ031D)

# Reading in self-reported health status (covariate)
health <- read_xpt("https://wwwn.cdc.gov/Nchs/Nhanes/2017-2018/HSQ_J.XPT") %>%
  select(SEQN, HSD010)

# Reading in disability data (covariate)
disab <- read_xpt("https://wwwn.cdc.gov/Nchs/Nhanes/2017-2018/DLQ_J.XPT") %>%
  select(SEQN, DLQ010, DLQ020, DLQ040, DLQ050, DLQ060, DLQ080)

# Reading in self-reported oral health status and utilization data (outcome)
ohques <- read_xpt("https://wwwn.cdc.gov/Nchs/Nhanes/2017-2018/OHQ_J.XPT") %>%
  select(SEQN, OHQ030, OHQ845)

# Reading in examiner-assessed oral health status data (outcome)
ohexam <- read_xpt("https://wwwn.cdc.gov/Nchs/Nhanes/2017-2018/OHXREF_J.XPT") %>%
  select(SEQN, OHAREC, OHDEXSTS)
```

Previewing data and creating merged data frame.
```{r}
# Previewing data
head(demo)
head(insure)
head(health)
head(disab)
head(ohques)
head(ohexam)

# Joining datasets by SEQN (respondent sequence number)
nhanesoh <- left_join(demo, insure, by = "SEQN")
nhanesoh <- left_join (nhanesoh, health, by = "SEQN")
nhanesoh <- left_join (nhanesoh, disab, by = "SEQN")
nhanesoh <- left_join (nhanesoh, ohques, by = "SEQN")
nhanesoh <- left_join (nhanesoh, ohexam, by = "SEQN")
```

```{r}
# Getting an overview of the data
names(nhanesoh)
summary(nhanesoh)
```

Defining covariates (age, gender, race/ethnicity, education, health insurance, self-reported health status, disability).
Defining outcomes (oral health utilization, self-reported oral health status, examiner-assessed oral health status)

 "HIQ011"   "HIQ031A" 
[13] "HIQ031B"  "HIQ031D"  "HSD010"   "DLQ010"   "DLQ020"   "DLQ040"   "DLQ050"   "DLQ060"   "DLQ080"   "OHQ030"   "OHQ845"   "OHAREC"  
[25] "OHDEXSTS"
```{r}
# Keeping gender, race/ethnicity, education, self-reported health status "as-is" for now.

# Creating categorical variable for age
nhanesoh <- mutate(nhanesoh, RIDAGEYR, agecat =
    ifelse(RIDAGEYR %in% 0:19, "0-19",
    ifelse(RIDAGEYR %in% 20:29, "20-29",
    ifelse(RIDAGEYR %in% 30:39, "30-39",
    ifelse(RIDAGEYR %in% 40:49, "40-49",
    ifelse(RIDAGEYR %in% 50:59, "50-59",
    ifelse(RIDAGEYR %in% 60:69, "60-69",
    ifelse(RIDAGEYR %in% 70:79, "70-79",
    ifelse(RIDAGEYR %in% 80:89, "80-89",
    ifelse(RIDAGEYR %in% 90:99, "90-99", "unsure"
    ))))))))))

# Defining disability covariate as dichotomous (presence of a disability = 1, no disability = 0)
nhanesoh <- mutate(nhanesoh, disability = ifelse(DLQ010==1 | DLQ020==1 | DLQ040==1 | DLQ050==1 | DLQ060==1 | DLQ080==1, 1, 0))

# Creating categorical variable for insurance
nhanesoh <- mutate(nhanesoh,
    healthins = case_when(
    HIQ011==2 ~ 0, #uninsured
    HIQ031D==17 ~ 1, #medicaid
    HIQ031A==14 ~ 2, #privateins
    HIQ031B==15 ~ 3 #medicare
    ))

summary(nhanesoh)
```

Refining data frame with relevant covariates, exposure, and outcomes of interest.
```{r}
# Removing irrelevant data frames
remove(demo, disab, health, insure, ohexam, ohques)

# Checking how many in sample had complete dental exam
table(nhanesoh$RIDSTATR)

# Filtering on individuals who were both interviewed and examined (RIDSTATR), filtering on individuals who received complete dental exam (OHDEXSTS) 
# From NHANES documentation, there are no missing values for gender, age, race/ethnicity
nhanesoh <- nhanesoh %>%
  filter(RIDSTATR==2) %>% #Dropped 550 observations (if not both interviewed adn examined)
  filter(OHDEXSTS==1) %>% #Dropped 267 observations with partial or not done oral health exams
  drop_na(disability) %>% #Dropped 2159 observations with missing information on disability
  #drop_na(education) %>% #Dropped 884 observations with missing information on education
  drop_na(HSD010) %>% #Dropped 267 observations with missing information on self-reported health status
# At this point, no missing values for ohutil, srohstatus, exohstatus
# Updating data frame with created exposure and outcome variables, covariates (renamed), and relevant survey weights
  select(SEQN, RIAGENDR, RIDAGEYR, RIDRETH3, DMDEDUC2,
  WTINT2YR, WTMEC2YR, SDMVPSU, SDMVSTRA, HSD010, disability, OHQ030, OHQ845, OHAREC, healthins, agecat) %>%
  rename(
    seqn = SEQN,
    gender = RIAGENDR,
    age = RIDAGEYR,
    "race/ethnicity" = RIDRETH3,
    education = DMDEDUC2,
    srhealthstatus = HSD010,
    psu = SDMVPSU,
    strata = SDMVSTRA,
    interviewweight = WTINT2YR,
    examweight = WTMEC2YR,
    srohstatus = OHQ845,
    exohstatus = OHAREC,
    ohutil = OHQ030
  )

names(nhanesoh)

# Recoding education, privateins, srohstatus, gender, uninsured, srhealthstatus, exohstatus, medicaid, disability, race/ethnicity, medicare, ohutil to factor variables from numeric
nhanesoh$education <- as.factor(nhanesoh$education)
nhanesoh$srohstatus <- as.factor(nhanesoh$srohstatus)
nhanesoh$gender <- as.factor(nhanesoh$gender)
nhanesoh$srohstatus <- as.factor(nhanesoh$srohstatus)
nhanesoh$exohstatus <- as.factor(nhanesoh$exohstatus)
nhanesoh$disability <- as.factor(nhanesoh$disability)
nhanesoh$"race/ethnicity" <- as.factor(nhanesoh$"race/ethnicity")
nhanesoh$ohutil <- as.factor(nhanesoh$ohutil)

head(nhanesoh)
summary(nhanesoh)

# Further cleaning dataset
nhanesoh <- nhanesoh %>%
  filter(education %in% 1:5) %>%
  filter(srhealthstatus %in% 1:5) %>%
  filter(ohutil %in% 1:7) %>%
  filter(srohstatus %in% 1:5) %>%
  filter(exohstatus %in% 1:4)
```

```{r}
# According to CDC tutorial (https://wwwn.cdc.gov/nchs/nhanes/tutorials/Module3.aspx), since I am using both interview and exam data, I should use the MEC exam weight

# Creating variable "nhanesdesign" to use in runnning analysis, using "svydesign" to assign weights
nhanesdesign <- svydesign(id      = ~psu,
                          strata  = ~strata,
                          weights = ~examweight,
                          nest    = TRUE,
                          data    = nhanesoh)

# Applying weights to covariates (age, agecat, gender, race/ethnicity, education, srhealthstatus, disability, healthins)
agedesign <- subset(nhanesdesign, age %in% 0:200)
svymean(~age, agedesign, na.rm = TRUE)
summary(nhanesoh$age)

# Applying weights to outcomes (srohstatus, exohstatus, ohutil)

```

library(ggplot2)

ggplot(nhanesoh) +
 aes(x = ohutil, fill = gender) +
 geom_bar() +
 scale_fill_hue() +
 theme_minimal() +
 facet_wrap(vars(disability))

ggplot(nhanesoh) +
 aes(x = ohutil, fill = `race/ethnicity`) +
 geom_bar() +
 scale_fill_hue() +
 theme_minimal() +
 facet_wrap(vars(disability))

ggplot(nhanesoh) +
 aes(x = ohutil, fill = education) +
 geom_bar() +
 scale_fill_hue() +
 theme_minimal() +
 facet_wrap(vars(disability))

ggplot(nhanesoh) +
 aes(x = ohutil, fill = education) +
 geom_bar() +
 scale_fill_hue() +
 theme_minimal() +
 facet_wrap(vars(disability))

ggplot(nhanesoh) +
 aes(x = ohutil, fill = agecat) +
 geom_bar() +
 scale_fill_hue() +
 theme_minimal() +
 facet_wrap(vars(disability))



```{r}
esquisse::esquisser()
```




```{r}
ggplot(nhanesoh) +
 aes(x = ohutil) +
 geom_bar(fill = "#0c4c8a") +
 theme_minimal()
```





























```{r}
ggplot(data = nhanes) +
   geom_mosaic(aes(x = product(ohutil), fill=ohutil, conds=product(disability), na.rm=TRUE))
```


```{r}
# Adjusted models
#logistic <- svyglm(ohutil~age+gender+education+disability+srhealthstatus, design=as.svrepdesign(nhanesdesign), family=quasibinomial, return.replicates=TRUE)
#logistic

#https://cran.r-project.org/web/packages/ggmosaic/vignettes/ggmosaic.html

#logistic <- svyglm(ohutil~+++srhealthstatus+uninsured+privateins+medicaid+medicare+disability, #design=as.svrepdesign(nhanesdesign), family=quasibinomial, return.replicates=TRUE)

```


ADDITIONAL CODE IDEAS (NOT USED)


```{r}



# Defining outcome (oral health utilization) as dichotomous (no utilization of dental services in past year = 1, utilization in past year = 0)
#nhanes <- mutate(nhanes, ohutil = ifelse(OHQ030==3 | OHQ030==4 | OHQ030==5, 1, 0))

# Defining outcome (self-reported oral health status) as dichotomous (fair/poor self-reported oral health = 1, excellent/good/very good self-reported oral health = 0)
#nhanes <- mutate(nhanes, srohstatus = ifelse(OHQ845==4 | OHQ845==5, 1, 0))

# Defining outcome (assessed oral health status by examiner) as dichotomous (dental care needed, proxy of fair/poor oral health status = 1, dental care not needed, proxy of excellent/good/very good oral health status = 0)
#nhanes <- mutate(nhanes, exohstatus = ifelse(OHAREC==1 | OHAREC==2 | OHAREC==3, 1, 0))
library(ggplot2)

ggplot(nhanesoh) +
 aes(x = ohutil) +
 geom_bar(fill = "#0c4c8a") +
 theme_minimal()


# Creating education indicator variables (education level for adults 20+) (high less than high school = 1, high school or greater = 0)
#nhanes <- mutate(nhanes, education = ifelse(DMDEDUC2==1 | DMDEDUC2==2, 1, 0))


# Creating indicator variable for "uninsured"
#nhanes <- mutate(nhanes, uninsured = ifelse(HIQ011==2, 1, 0))
# Creating indicator variable for "covered by Medicaid"
#nhanes <- mutate(nhanes, medicaid = ifelse(HIQ031D==17, 1, 0))
## Creating indicator variable for "covered by Medicare"
#nhanes <- mutate(nhanes, medicare = ifelse(HIQ031B==15, 1, 0))
# Creating indicator variable for "covered by private insurance"
#nhanes <- mutate(nhanes, privateins = ifelse(HIQ031A==14, 1, 0))

# Creating categorical variable for health insurance
#nhanes <- mutate(nhanes, healthins =
    #ifelse(uninsured==1, "uninsured",
    #ifelse(medicaid==1, "medicaid",
    #ifelse(medicare==1, "medicare",
    #ifelse(privateins==1, "private"
    #)))))
```
